---
title: "Process scRNA-seq"
---
##Location of script
```{r}
here::i_am("Code/Preprocess_scRNA.Rmd")
```
##Download Data
We import the scRNA data from (Tepe et al, 2018),
Under accession number GSE121891
We have two sets of meta_data.
With two levels of annotation 
We use the annotation the same as the CARD paper 
```{r}
MOBSC_sce_raw <- read.csv(gzfile(here("Data/scrna_mob/Raw","GSE121891_OB_6_runs.raw.dge.csv.gz")))

meta_data <- read.csv(gzfile(here("Data/scrna_mob/Meta","GSE121891_OB_metaData_seurat.csv.gz")))

meta_data2 <- read.csv(here("Data/scrna_mob/Meta", "GSE121891_Figure_2_metadata.txt"), sep = "\t")

MOBSC_sce <- readRDS((url("https://figshare.com/ndownloader/files/40581983")))

```
##Create Seurat Object 
###Add Metadata 
```{r}
rownames(MOBSC_sce_raw) <- MOBSC_sce_raw[[1]]  # Set the first column as row names


seurat_object1 <- CreateSeuratObject(counts = MOBSC_sce_raw)



seurat_object2 <- as.Seurat(MOBSC_sce) ## 182 genes 


seurat_object1 <- subset(seurat_object1, cells = colnames(seurat_object2))

# Extract cell names from both objects
cellnames_object1 <- rownames(seurat_object1@meta.data)
cellnames_object2 <- rownames(seurat_object2@meta.data)

# Check if the cell names match (you can use set operations to find mismatches)
mismatched_cells <- setdiff(cellnames_object1, cellnames_object2)

celltypes_data <- seurat_object2@meta.data$cell_type

# Add the 'celltypes' column to seurat_object1's metadata
seurat_object1@meta.data$celltypes <- celltypes_data[match(cellnames_object1, cellnames_object2)]

# Add FinalIds to meta_data
seurat_object1@meta.data$celltypes2 <- meta_data2$FinalIds[match(rownames(seurat_object1@meta.data), rownames(meta_data2))]

# Add ClusterName column to seurat_object1's metadata
##After subset only neuronal cell types are selected 
seurat_object1@meta.data$celltypes3 <- meta_data$ClusterName[match(rownames(seurat_object1@meta.data), meta_data$X)]
```
##Preprocessing 
```{r}
genes_to_exclude <- c("Malat1",mt_genes)

# Invert the exclusion to get a list of genes to keep
genes_to_keep <- setdiff(rownames(seurat_object1), genes_to_exclude)

# Subset the Seurat object to keep only the desired genes
seurat_object1 <- subset(seurat_object1, features = genes_to_keep)

seurat_object1[["percent.mt"]] <- PercentageFeatureSet(seurat_object1, pattern = "^mt-")

seurat_object1<- subset(seurat_object1, subset = nFeature_RNA > 200 & percent.mt < 60)

##Filtering for genes with expression above zero, rowsums of counts data should be above zero 
genes_with_expression <- rownames(seurat_object1)[rowSums(GetAssayData(seurat_object1, slot = "counts")) > 0]

seurat_object1 <- subset(seurat_object1, features = genes_with_expression)


###Normalize and scale####
seurat_object1 <- NormalizeData(seurat_object1, normalization.method = "LogNormalize", scale.factor = 10000)

seurat_object1 <- ScaleData(seurat_object1, features = rownames(seurat_object1))
# Run PCA
seurat_object1 <- FindVariableFeatures(seurat_object1, selection.method = "vst", nfeatures = 2000)
#
top10 <- head(VariableFeatures(seurat_object1, ),10)

plot1 <- VariableFeaturePlot(seurat_object1)
LabelPoints(plot = plot1, points = top10, repel = TRUE, xnudge = 0, ynudge = 0)

# Run PCA
seurat_object1 <- RunPCA(seurat_object1, features = VariableFeatures(object = seurat_object1))
```
##EDA
```{r}
seurat_object1 <- RunUMAP(seurat_object1 , dims = 1:15)
 ##Celltypes 2 contains higher level of cell type annotations
p1 <- DimPlot(seurat_object1 , group.by = "celltypes2", label = TRUE, reduction = "umap" )
p2 <- DimPlot(seurat_object1, group.by = "celltypes",label = TRUE, reduction = "umap"  )
p1 + p2 

```
##Subset scRNA and ST
(Run Preprocess_ST.Rmd first)
```{r}
seurat_object1 <- subset(seurat_object1, features = rownames(spe))
```
##Create Filtered Gene Sets
###No Ribosomal Genes
```{r}
Rp_genes <- grep("^Rp", rownames(seurat_object1), value = TRUE)
genes_to_exclude <- c(Rp_genes)
genes_to_keep <- setdiff(rownames(seurat_object1), genes_to_exclude)
seurat_object1_norp <- subset(seurat_object1,features = genes_to_keep)
```
###No platform bias genes
(Run Platform_bias.R first)
```{r}
significantly_diff_genes 
genes_to_keep <- setdiff(rownames(seurat_object1), significantly_diff_genes)
seurat_object1_sig_diff <- subset(seurat_object1,features = genes_to_keep)
```
##Function for finding marker genes
```{r}
get_top_markers <- function(markers_all, total_genes) {
  # Calculate the number of genes per cell type
  genes_per_type <- total_genes %/% 4
  
  # Filter and sort markers for each cell type
  sorted_markers_GC <- markers_all %>%
    filter(cluster == 'GC', avg_log2FC >= 0) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  sorted_markers_PGC <- markers_all %>%
    filter(cluster == 'PGC', avg_log2FC >= 0) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  sorted_markers_OSNS <- markers_all %>%
    filter(cluster == 'OSNs', avg_log2FC >= 0) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  sorted_markers_MTC <- markers_all %>%
    filter(cluster == 'M/TC', avg_log2FC >= 0) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  # Extract gene names
  genes_GC <- rownames(sorted_markers_GC)
  print(length(genes_GC))
  genes_PGC <- rownames(sorted_markers_PGC)
  print(length(genes_PGC))
  genes_OSNS <- rownames(sorted_markers_OSNS)
  print(length(genes_OSNS))
  genes_MTC <- rownames(sorted_markers_MTC)
  print(length(genes_MTC))
  
  # Combine all top markers into one list
  print(Reduce(intersect, list(genes_GC, genes_PGC, genes_OSNS, genes_MTC)))
  deg <- unique(c(genes_GC, genes_PGC, genes_OSNS, genes_MTC))
  
  # Return the result
  return(deg)
}
```
##Function for finding marker genes with cutoff
```{r}
get_top_markersFC_cutoff <- function(markers_all, total_genes, cutoff = 1.5) {
  # Calculate the number of genes per cell type
  genes_per_type <- total_genes %/% 4
  
  # Filter and sort markers for each cell type
  sorted_markers_GC <- markers_all %>%
    filter(cluster == 'GC', avg_log2FC >= cutoff) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  sorted_markers_PGC <- markers_all %>%
    filter(cluster == 'PGC', avg_log2FC >= cutoff) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  sorted_markers_OSNS <- markers_all %>%
    filter(cluster == 'OSNs', avg_log2FC >= cutoff) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)
  
  sorted_markers_MTC <- markers_all %>%
    filter(cluster == 'M/TC', avg_log2FC >= cutoff) %>%
    arrange(p_val_adj) %>%
    head(genes_per_type)

  
  # Extract gene names
  genes_GC <- rownames(sorted_markers_GC)
  print(length(genes_GC))
  genes_PGC <- rownames(sorted_markers_PGC)
  print(length(genes_PGC))
  genes_OSNS <- rownames(sorted_markers_OSNS)
  print(length(genes_OSNS))
  genes_MTC <- rownames(sorted_markers_MTC)
  print(length(genes_MTC))
  
  # Combine all top markers into one list
  print(Reduce(intersect, list(genes_GC, genes_PGC, genes_OSNS, genes_MTC)))
  deg <- unique(c(genes_GC, genes_PGC, genes_OSNS, genes_MTC))
  
  
  markers <- markers_all[markers_all$cluster == 'GC', ]
  markersPGC <- markers_all[markers_all$cluster == 'PGC', ]
  markersOSNS <- markers_all[markers_all$cluster == 'OSNs', ]
  markersMTC <- markers_all[markers_all$cluster == 'M/TC', ]
  
  create_volcano_plot <- function(data, log2fc_cutoff, pval_cutoff, dataset_name, n_labels = 10) {
    
    # Add new columns to the data
    data <- data %>%
      mutate(logP = -log10(p_val_adj),
             Significant = ifelse(p_val_adj < pval_cutoff & abs(avg_log2FC) > log2fc_cutoff, "Yes", "No"),
             is_Rp = grepl("^Rp", gene)) # Identify Rp genes
    
    # Identify top significant genes
    top_genes <- data %>%
      filter(Significant == "Yes") %>%
      arrange(desc(logP), desc(abs(avg_log2FC))) %>%
      slice_head(n = n_labels) %>%
      mutate(rank = row_number()) # Add ranking numbers
    
    # Modify color mapping
    data <- data %>%
      mutate(Color = case_when(
        Significant == "Yes" & is_Rp == TRUE ~ "Rp",
        Significant == "Yes" & is_Rp == FALSE ~ "Significant",
        TRUE ~ "Non-Significant"
      ))
    
    # Create the volcano plot
    p <- ggplot(data, aes(x = avg_log2FC, y = logP, color = Color)) +
      geom_point(alpha = 0.5) +
      scale_color_manual(values = c("Non-Significant" = "grey", "Significant" = "red", "Rp" = "green")) +
      geom_text(data = top_genes, aes(label = paste(rank, gene, sep = ". ")), vjust = 2, size = 3, check_overlap = TRUE, color = "blue") + # Add ranking numbers to gene labels
      theme_minimal() +
      labs(title = paste("Volcano Plot for", dataset_name), x = "Log2 Fold Change", y = "-Log10 Adjusted p-value") +
      geom_vline(xintercept = c(-log2fc_cutoff, log2fc_cutoff), color = "blue", linetype = "dashed") +
      geom_hline(yintercept = -log10(pval_cutoff), color = "blue", linetype = "dashed")
    
    return(p)
  }
  
  gc_volcano_plot <- create_volcano_plot(markers, log2fc_cutoff = cutoff, pval_cutoff = 0.01, dataset_name = "GC", n_labels = 55)
  PGC_volcano_plot <- create_volcano_plot(markersPGC, log2fc_cutoff = cutoff, pval_cutoff = 0.01, dataset_name = "PGC", n_labels = 55)
  OSN_volcano_plot <- create_volcano_plot(markersOSNS, log2fc_cutoff = cutoff, pval_cutoff = 0.01, dataset_name = "OSNs", n_labels = 55)
  MTC_volcano_plot <- create_volcano_plot(markersMTC, log2fc_cutoff = cutoff, pval_cutoff = 0.01, dataset_name = "M/TC", n_labels = 55)
  
  plots <- gc_volcano_plot | PGC_volcano_plot |  OSN_volcano_plot | MTC_volcano_plot
  
  
  # Return the result
  return(deg)
}
```
##Create Gene Sets
```{r}
deg200 <- get_top_markers(markers_all, 220)
deg400 <- get_top_markers(markers_all, 440)
deg600 <- get_top_markers(markers_all, 680)
deg_norp <-  get_top_markers(markers_all_norp, 220)
deg_cutoff1.5 <- get_top_markersFC_cutoff(markers_all, 220)
deg_sigdiff <- get_top_markers(markers_all_sigdiff, 220)
```
##Process and save dataset for BLADE
(To create expression profiles required by BLADE)
```{r}
process_and_save_geneset <- function(seurat_object, geneset, geneset_name, base_dir = here("Data", "Processed_BLADE")) {
  # Subset the Seurat object
  seurat_object_filtered <- subset(seurat_object, features = geneset)
  
  # Define helper functions
  calc_std_dev <- function(seurat_obj) {
    data <- GetAssayData(seurat_obj, assay = "RNA", slot = "data")
    return(rowSds(as.matrix(data)))
  }
  
  calc_mean_expr <- function(seurat_obj) {
    data <- GetAssayData(seurat_obj, assay = "RNA", slot = "data")
    return(rowMeans(as.matrix(data)))
  }
  
  calc_variance <- function(seurat_obj) {
    data <- GetAssayData(seurat_obj, assay = "RNA", slot = "data")
    return(rowVars(as.matrix(data)))
  }
  
  calc_corrected_std <- function(mean_matrix, variance_matrix) {
    gene_list <- rownames(mean_matrix)
    ct_list <- colnames(mean_matrix)
    New_std <- matrix(nrow = length(gene_list), ncol = length(ct_list))
    
    for (i in seq_along(ct_list)) {
      trend <- fitTrendVar(mean_matrix[, i], variance_matrix[, i])$trend
      New_std[, i] <- sqrt(trend(mean_matrix[, i]))
    }
    
    rownames(New_std) <- gene_list
    colnames(New_std) <- ct_list
    return(New_std)
  }
  
  # Split the Seurat object by cell type
  seurat_list <- SplitObject(seurat_object_filtered, split.by = "celltypes")
  
  # Apply the function to each cell type subset
  std_dev_list <- lapply(seurat_list, calc_std_dev)
  std_dev_matrix <- do.call(cbind, std_dev_list)
  colnames(std_dev_matrix) <- names(std_dev_list)
  
  mean_expr_list <- lapply(seurat_list, calc_mean_expr)
  mean_expr_matrix <- do.call(cbind, mean_expr_list)
  colnames(mean_expr_matrix) <- names(mean_expr_list)
  
  variance_list <- lapply(seurat_list, calc_variance)
  variance_matrix <- do.call(cbind, variance_list)
  colnames(variance_matrix) <- names(variance_list)
  
  gene_names <- rownames(GetAssayData(seurat_object_filtered, assay = "RNA", slot = "data"))
  rownames(std_dev_matrix) <- gene_names
  rownames(mean_expr_matrix) <- gene_names
  rownames(variance_matrix) <- gene_names
  
  New_std <- calc_corrected_std(mean_expr_matrix, variance_matrix)
  
  # Save the datasets
  datasets <- list(
    mean_expr_matrix = paste0("deg_mean", geneset_name, ".csv"),
    New_std = paste0("deg_sd", geneset_name, ".csv")
  )
  
  base_dir_full <- here(base_dir)
  
  for (data_name in names(datasets)) {
    file_path <- file.path(base_dir_full, datasets[[data_name]])
    
    if (!dir.exists(dirname(file_path))) {
      dir.create(dirname(file_path), recursive = TRUE)
    }
    
    write.csv(get(data_name), file = file_path, row.names = TRUE)
    message(paste("Saved", datasets[[data_name]]))
  }
  
  return(list(mean_expr_matrix = mean_expr_matrix, New_std = New_std))
}
```

```{r}
results200 <- process_and_save_geneset(seurat_object1, deg200, "200")
results400 <- process_and_save_geneset(seurat_object1, deg400, "400")
results600 <- process_and_save_geneset(seurat_object1, deg600, "600")
results_norp <- process_and_save_geneset(seurat_object1_norp, deg_norp, "_norp")
results_sigdiff <-  process_and_save_geneset(seurat_object1_sig_diff, deg_sigdiff, "_sigdiff200")
results_cutoff <-  process_and_save_geneset(seurat_object1, deg_cutoff1.5, "_cutoff1.5")
```

